Search-based Synthesis of Network Intrusion Detection Rules

### 2019-09-16 paper/paper.tex
pass on technique

\tname{} is a search-based approach that uses positive and negative
network traffic examples to synthesize signature-based NIDS rules,
such as those from Snort and Suricata. \tname{} synthesizes rules in
two steps. First, it produces a potentially overspecified rule that
captures the negative example that is provided as input. Then, it uses
positive examples to eliminate unecessary options from the rule. The
goal is to create a rule that captures only the negative traffic.

@@

One important step in search-based optimization is to characterize a
candidate solution to the target problem. Candidate solutions are the
objects that the search tries to optimize. In this case, a rule is a
candidate solution.  \tname\ encodes the rule options segment of a
Suricata rule (see Section~\ref{sec:example-suricata-rules}) as a
vector of Option objects. Some options are discarded as they only
serve for documentation. For example, the options \CodeIn{msg},
\CodeIn{classtype}, and \CodeIn{miscactivity} have no influence in
triggering the action of rules. The option \CodeIn{flags}, appearing
in the example on Figure~\ref{fig:synflood-example}, is relevant to
trigger the action. \tname{} models that option with a data structure
storing two fields, one of character type and another of integer
type. This enables \tname{} to mutate each element
separately.

@@

\pgfplotsset{width=6cm,compat=1.8}
\begin{figure}[h!]
  \centering
  \begin{tikzpicture}
    \begin{axis}[
        ybar stacked,
        enlargelimits=0.15,
        legend style={at={(0.5,-0.15)},
          anchor=north,legend columns=-1},
        ylabel={\#rules},
        symbolic x coords={gen, http, icmp, udp, tcp},
        xtick=data,
        nodes near coords,
        nodes near coords align={vertical},
      ]
      \addplot coordinates {(gen,70) (http,34) (icmp,94) (udp,84) (tcp,43)};
      \addplot coordinates {(gen,50) (http,20) (icmp,134) (udp,90) (tcp,123)};    
      \legend{relevant, irrelevant}
    \end{axis}
  \end{tikzpicture}
  \caption{\label{fig:distribution-rules-protocol}Distribution of
    rules per protocol and relevance for triggering action.}
\end{figure}

@@

The fitness value of a candidate solution $c$, modelling the rule
options $o_1, \dots, o_n$ is given by $f(c)=1-\Sigma{d(o_i,m)}/n$,
where $i$ ranges from 1 to $n$ and $m$ denotes the negative
traffic. The higher the fitness value the better, \ie{}, the fitter
the candidate solution $c$ will be.  The value of the expression
$d(o_i,m)$, associated with option $o_i$, ranges over
0-1. Consequently, the value of the expression $\Sigma{d(o_i,m)}/n$
itself is in the 0-1 range, as there are $n$ options, and the range of
the function is 0-1.

As usual, we used customized distance functions to compute $d(o_i,
m)$. For example, let us consider the \CodeIn{flags} option, discussed
on Figure~\ref{fig:synflood-example}. Recall that this option
identifies the SYN packet in the tcp protocol and that its encoding in
\tname{} stores two attributes---one categorical and one
numerical. The value of $d(o_i, m)$, in this case, is obtained by
computing the normalized Euclidian distance in the two-dimensional
space. Considering that each of the two dimensions has cardinality of
ten, the distance of a candidate solution holding the option
\CodeIn{flags(S,13)} to the solution will be 0.22
(=$1/\sqrt{20}$). Note that the smaller the distance the better.

@@

%%  LocalWords:  Suricata SolarWinds QRadar NIDS OISF TCP ACK

>>>>

\tname{} is a search-based approach that uses positive and negative
network traffic examples to synthesize signature-based NIDS rules,
such as those from Snort and Suricata. \tname{} synthesizes rules in
two steps. First, it produces a potentially overspecified rule that
captures the negative example that is provided as input. Then, it uses
positive examples to eliminate unnecessary options from the rule. The
goal of \tname\ is to create a rule that captures only the negative traffic.

@@

One important step in search-based optimization is to characterize a
candidate solution to the target problem. Candidate solutions are the
objects that the search optimize towards solutions. In this case, a a
candidate solution is an encoding of a Suricata rule.  \tname\ encodes
the rule options segment of a Suricata rule (see
Section~\ref{sec:example-suricata-rules}) as a vector of Option
objects. Some options are discarded as they only serve for
documentation. For example, the options \CodeIn{msg},
\CodeIn{classtype}, and \CodeIn{misc-activity}, appearing in the
example on Figure~\ref{fig:synflood-example}, have no influence in
triggering the actions associated with rules. The option
\CodeIn{flags}, however, is relevant to trigger the action. \tname{}
models that option with a data structure storing two fields, one of
character type and another of integer type. This enables \tname{} to
mutate each element separately. Similar encodings are used in modeling
the other relevant options. We found what is (ir)relevant by
inspecting Suricata code for the presence of \CodeIn{eval} functions,
which are invoked from a main look to process every Suricata rule. The
presence of such a function for a given option indicates that it is
relevant for triggering some action of some rule.

@@

\pgfplotsset{width=5.5cm,compat=1.8}
\begin{figure}[h!]
  \centering
  \begin{tikzpicture}
    \begin{axis}[
        ybar stacked,
        enlargelimits=0.15,
        legend style={at={(0.5,-0.15)},
          anchor=north,legend columns=-1},
        ylabel={\#rules},
        symbolic x coords={gen, http, icmp, udp, tcp},
        xtick=data,
        nodes near coords,
        nodes near coords align={vertical},
      ]
      \addplot coordinates {(gen,70) (http,34) (icmp,94) (udp,84) (tcp,43)};
      \addplot coordinates {(gen,50) (http,20) (icmp,134) (udp,90) (tcp,123)};    
      \legend{relevant, irrelevant}
    \end{axis}
  \end{tikzpicture}
  \caption{\label{fig:distribution-rules-protocol}Distribution of
    rules per protocol and relevance for triggering action.}
\end{figure}

\subsection{Initial State}

@@

The fitness value of a candidate solution $c$, modeling the rule
options $o_1, \dots, o_n$ is given by $f(c)=1-\Sigma{d(o_i,m)}/n$,
where $i$ ranges from 1 to $n$ and $m$ denotes the negative
traffic. The higher the fitness value the better, \ie{}, the fitter
the candidate solution $c$ will be.  The value of the expression
$d(o_i,m)$, associated with option $o_i$, ranges over
0-1. Consequently, the value of the expression $\Sigma{d(o_i,m)}/n$
itself is in the 0-1 range, as there are $n$ options, and the range of
the function is 0-1.

As usual, we used customized distance functions to compute $d(o_i,
m)$. For example, let us consider the \CodeIn{flags} option, discussed
on Figure~\ref{fig:synflood-example}. Recall that this option
identifies the SYN packet in the tcp protocol and that its encoding in
\tname{} stores two attributes---one categorical and one
numerical. The value of $d(o_i, m)$, in this case, is obtained by
computing the normalized Euclidean distance in the two-dimensional
space. Considering that each of the two dimensions has cardinality of
ten, the distance of a candidate solution holding the option
\CodeIn{flags(S,13)} to the solution will be 0.22
(=$1/\sqrt{20}$). Note that the smaller the distance the better.

@@

%%  LocalWords:  Suricata SolarWinds QRadar NIDS OISF TCP ACK msg tcp
%%  LocalWords:  encodings eval cardinality
